---
title: "GAMs 101"
output: 
  html_notebook:
    toc: true 
    fig_height: 3
    fig_width: 4
---
******************************
##### Dataset

This notebook aims at presenting basics of Generalized Additive Models (Hastie & Tibsharani, 1996). We are agoing to focus on fictitious (though probable) data of a subscription business where people sign up for receiving meal boxes. 

```{r, results='hide'}
library.list <- c('mgcv', 'forecast', 'openxlsx', 'ggplot2', 'e1071', 'dplyr', 'tidyr')
lapply(library.list, require, character.only=T)
```

```{r, results='markup'}
set.seed = 2018
setwd('./data/')
df <- read.xlsx('subscription_demand.xlsx')
df$year.num <- c(rep(1, 52), rep(2, nrow(df)-52))
df$year <- as.factor(paste0('Year', df$year.num))
head(df)
summary(df)
```

************************

Our dataset consists of **5 predictors** and a target ('sales') in form of number of boxes to be shipped each week. Our 'planned' predictors are measured 4 weeks in advance, simply as number of people who plan to churn, pause their subscription or receive a box as usual. Additionally, we will be looking at:
- **holiday effect**: note this is a continuous and normalized variable, aggregting effects of seasonality and one-time events
- **trend**: a variable denoting 'baseline' trend of sales, computed on all data available until t-1 for moment t0; also normalized

Interestingly, **no clear time-related effects** can easily be spotted within our target which suggests that forecasting based on traditional time-series methods (like ARIMAx) might not prove to be the best option.

```{r}
acf(diff(df$sales))
pacf(diff(df$sales))
```

```{r, results='hide'}
df$sales_lag3 <- c(rep(df$sales[1],3), df$sales[1:(nrow(df)-3)])
plot(df$sales, type='l')
lines(df$sales_lag3, col='blue')
```

##### Basic GAM

Let's start with a simple split into train & test. On train set we'll build a basic GAM, letting the package choose number of knots and going with default smooth settings ('tp' - thin plate splines).

```{r, results='markup'}
df.train <- df[1:87,]
demand.gam.model <- gam(sales ~ s(holiday.effect) + s(planned_sales) + 
                s(planned_pauses) + s(planned_churn) +
                + s(trend), data = df.train, control=gam.control(mgcv.tol=1e-4))
summary(demand.gam.model)
```


```{r}
plot(demand.gam.model, pages = 1, shade=T, residuals = T)
```

##### Smooths & regularization

A large set of smooth options is available. Most common include:
  - thin plate splines ('tp')
  - B-splines ('bs')
  - a subset of B-splines: cubic regression splines ('cr' or shrunk version, 'cs')

See what happens if we switch to cubic splines:

```{r}
demand.gam.model <- gam(sales ~ s(holiday.effect, bs='cr') + s(planned_sales, bs='cr') + 
                s(planned_pauses, bs='cr') + s(planned_churn, bs='cr') +
                + s(trend, bs='cr'), data = df.train, control=gam.control(mgcv.tol=1e-4))
summary(demand.gam.model)

```

```{r}
plot(demand.gam.model, pages = 1, shade=T, residuals = T)
```

We can use regularization, either via a penalty applied to all terms with select=T or using a shrinkage version of smooths.

```{r}
demand.gam.model <- gam(sales ~ s(holiday.effect, bs='cr') + s(planned_sales, bs='cr') + 
                s(planned_pauses, bs='cr') + s(planned_churn, bs='cr') +
                + s(trend, bs='cr'), data = df.train, control=gam.control(mgcv.tol=1e-4), select=T)
summary(demand.gam.model)
```

```{r}
plot(demand.gam.model, pages = 1, shade=T, residuals = T)
```

```{r}
spline.type <- 'cs'
demand.gam.model <- gam(sales ~ s(holiday.effect, bs=spline.type) + s(planned_sales, bs=spline.type) + 
                s(planned_pauses, bs=spline.type) + s(planned_churn, bs=spline.type) +
                + s(trend, bs=spline.type), data = df.train, control=gam.control(mgcv.tol=1e-4))
summary(demand.gam.model)
```

If nature of our relation between predictors and target changes over time, we can introduce a 'by' variable.

```{r}
spline.type <- 'cs'

demand.gam.model <- gam(sales ~ year + s(holiday.effect, bs=spline.type, by=year) + s(planned_sales, bs=spline.type) + s(planned_pauses, bs=spline.type) + s(planned_churn, bs=spline.type) +
                + s(trend, bs=spline.type, by=year), data = df.train, control=gam.control(mgcv.tol=1e-4))
summary(demand.gam.model)
```

We can also restrict number of knots used by the algorithm for specific predictors.

```{r}
spline.type <- 'cs'
demand.gam.model <- gam(sales ~ year + s(holiday.effect, bs=spline.type, by=year) + s(planned_sales, bs=spline.type) + s(planned_pauses, bs=spline.type) + s(planned_churn, bs=spline.type, k = 3) +
                + s(trend, bs=spline.type), data = df.train, control=gam.control(mgcv.tol=1e-4))
summary(demand.gam.model)
```
```{r}
plot(demand.gam.model, pages = 1, shade=T, residuals = T)
```

Let's see if lagged sales could help or not...

```{r}
spline.type <- 'cs'
demand.gam.model <- gam(sales ~ year + s(holiday.effect, bs=spline.type, by=year) + s(planned_sales, bs=spline.type) + s(planned_pauses, bs=spline.type) + s(planned_churn, bs=spline.type) +
                + s(trend, bs=spline.type) +s(sales_lag3, bs=spline.type), data = df.train, control=gam.control(mgcv.tol=1e-4))
summary(demand.gam.model)
```

```{r}
plot(demand.gam.model, select=7, shade=T, residuals = T)
```

```{r, results='hide'}
param.grid <- expand.grid(cost = c(0.001, 0.01, 0.1, 1, 2, 3, 5, 10, 20, 50), gamma = c(0.01, 0.02, 0.05, 0.1, 0.25, 0.5, 1))

all.errors <- data.frame()
for (mm in (83:87)) {
  errors <- data.frame()
  for (p in 1:nrow(param.grid)) {
    svr.model <- svm(sales ~ ., data = df[1:mm,2:8], kernel = 'radial', cost = param.grid$cost[p],
                   gamma = param.grid$gamma[p])
    errors <- rbind(errors, c(param.grid$cost[p], param.grid$gamma[p], mean(abs(df[(mm+1):(mm+8),2] - predict(svr.model, newdata = df[(mm+1):(mm+8),2:8])))))
  }
  names(errors) <- c('cost','gamma',paste0('error', mm))
  if (mm==83) {
    all.errors <- errors
  } else {
    all.errors <- inner_join(all.errors, errors, by=c('cost','gamma'))
  }
}
all.errors$avg.error <- apply(all.errors[,3:7], 1, mean)
all.errors <- all.errors[order(all.errors$avg.error),]
all.errors[1:10,]
```

```{r, results='markup'}

all.errors <- data.frame()
max.obs <- 78:87
max.cv <- 8

for (m in max.obs) {
  final.gam <- gam(sales ~ s(holiday.effect, bs=spline.type) + s(planned_sales, bs=spline.type) + s(planned_pauses, bs=spline.type) + s(planned_churn, bs=spline.type) +
                + s(trend, bs=spline.type), data = df[1:m,], control=gam.control(mgcv.tol=1e-5))
 svr.model <- svm(sales ~ ., data = df[1:m,2:8], kernel = 'radial', cost = 5,
                   gamma = 0.02)
 arima.model <- auto.arima(df$sales[1:m], xreg=df[1:m,c('holiday.effect','year.num')])
 
 gam.preds <- predict(final.gam, newdata = df[(m+1):(m+max.cv),])
 svr.preds <- predict(svr.model, newdata = df[(m+1):(m+max.cv),])
 arima.preds <- as.numeric(forecast(arima.model, xreg=df[(m+1):(m+max.cv),c('holiday.effect','year.num')])$mean)
 gam.svr.preds <- .5 * (gam.preds + svr.preds)
 svr.arima.preds <- .5 * (svr.preds + arima.preds)
 gam.arima.preds <- .5 * (gam.preds + arima.preds)
 combined3preds <- (gam.preds + svr.preds + arima.preds) / 3
 
 actual.sales <- df$sales[(m+1):(m+max.cv)]
 pred.list <- list(gam.preds, svr.preds, arima.preds, gam.svr.preds, svr.arima.preds, gam.arima.preds,
                   combined3preds)
 errors <- sapply(pred.list, function(x) mean(abs(x - actual.sales)/actual.sales)) 
 all.errors <- rbind(all.errors, errors)
}
names(all.errors) <- c('gam','svr','arima','gam.svr','svr.arima','gam.arima','combined3preds')
all.errors$last.training.week <- paste0('W', max.obs) 
plot.df <- as.data.frame(all.errors %>% gather('method', 'MAPE', 1:7))

ggplot(aes(x=last.training.week, y = MAPE, group = method), data = plot.df) + 
  geom_line(aes(color=method)) + geom_point((aes(color=method))) + ylab(paste0('MAPE on next ',max.cv, ' weeks'))

```

```{r}
library(scam)
setwd('./data/')
marketing.data <- read.csv('marketing_dummy_data.csv')
head(marketing.data)
```

##### Marketing attribution

The key question is how to obtain contributions to target (eg conversions) from each channel where we spent money.

```{r}
mrktg.gam <- gam(Conversions ~ s(TVSpend, bs = spline.type) + s(Facebook, bs = spline.type) + 
            s(Affiliate, bs = spline.type) +
            s(SEM.nonbrand, bs = spline.type) + s(SEM.brand, bs = spline.type) + 
            Xmas.flag + s(Weekday, k = 7), data = marketing.data)
summary(mrktg.gam)
```

```{r}
df.terms <- predict(mrktg.gam, type='terms')
head(df.terms, 20)
plot(mrktg.gam, pages = 2, shade=T, residuals = T)
```

Some models require applying shape constraints so that we get non-negative impact of channels. Then new splines can be used, e.g:
- mpi - monotonic increasing penalized B-splines
- micv - monotonic convex increasing B-splines
etc.
Typically shape constraints result in underfitting .

```{r}
scam.spline.type = 'mpi'
#marketing.data$Weekday.factor <- as.factor(marketing.data$Weekday)
mrktg.scam <- scam(Conversions ~ s(TVSpend, bs = scam.spline.type) + s(Facebook, bs = scam.spline.type) + 
            s(Affiliate, bs = scam.spline.type) +
            s(SEM.nonbrand, bs = scam.spline.type) + s(SEM.brand, bs = scam.spline.type) + 
            Xmas.flag + s(Weekday, k=7), data = marketing.data)
summary(mrktg.scam)
```
```{r}
plot(mrktg.scam, pages=2, shade=T, residuals = T)
```


